---
title: "NFH temperature time series"
author: "Courtney Klepac"
date: "8/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##This is an R script that manipulates insitu HOBO temp data for descriptive statistics, ANoVA, correlations, and plots##
```{r}
#libraries
library(FSA)
library(dplyr)
library(lubridate)
library(tidyr)
library(zoo)
library(expss)
library(emmeans)
library(psych)
library(PerformanceAnalytics)
library(cowplot)
library(ggplot2)
```
##Import data and set factors and levels##
```{r}
setwd("~/Documents/MOTE/NFH")
getwd()

#read in hobotemp file
temp<-read.csv("NFHtemps.csv",header=T)
#temp<-read.delim(pipe("pbpaste"))
head(temp)
# to convert the DateTime column to and R DateTime format see ?strptime() for more explanation
temp$DateTime<-strptime(temp$DateTime, format="%m/%d/%y %I:%M %p")
#make month and year-month column to do summarize stats
temp$year=year(temp$DateTime)
temp$month=month(temp$DateTime)
temp$day<-as.numeric(day(temp$DateTime))
temp$Temp<-as.numeric(temp$Temp)
temp<-unite(temp, 'date', c('month','year'),remove=F)
temp<-unite(temp, 'd', c('day','month','year'),remove=F)
head(temp)
```
##daily descriptive stats##
```{r}
# use the day to calculate daily stats (range, min, max, mean)
tempdaily<-data.frame("dayRange"=tapply(temp$Temp, temp$d, function(x) range(x)[2]-range(x)[1]),"dayMin"=tapply(temp$Temp, temp$d, min),"dayMax"=tapply(temp$Temp, temp$d, max), "dayMean"=tapply(temp$Temp, temp$d, mean))
tempdaily$date<-row.names(tempdaily)
tempdaily$date<-strptime(tempdaily$date, format='%d_%m_%Y')
tempdaily<-tempdaily[order(tempdaily$date),]
head(tempdaily)
# print out summary stats for daily measurements
write.table(tempdaily,"NFHdailytempsummary.txt",sep = "\t")
```
##Boxplot of daily stats
```{r}
par(mfrow=c(1,3))
boxplot(tempdaily$dayRange, main = "Daily Range °C",ylab = "°C",notch=T, col="red",cex.lab=1.5,cex.axis=1.5)
boxplot(tempdaily$dayMean, main = "Daily Mean °C", ylab = "°C",notch=T, col="gold",cex.lab=1.5,cex.axis=1.75)
boxplot(tempdaily$dayMax, main = "Daily Max °C", ylab = "°C",notch=T, col="blue",cex.lab=1.5,cex.axis=1.8)
```
##plot of entire temperature trace
```{r}
#use the tickpos to set the axis labels you want to plot (note these need to match your first date-time file)
tickpos<-seq(as.POSIXct("2005-06-04"),as.POSIXct("2021-07-30"),by="1 year")
plot(tempdaily$date,tempdaily$dayMean, type="l", lwd= 1.5, ylab="Mean in-situ Temperature (°C)", xlab="Date", xaxt='n',ylim=c(18,34), col="blue")
# use points to add other temp traces to the same plot, will automatically use the date-time to plot them together
abline(h=29.7,lty=2,col="black")
legend("bottomright",c("NFH"),lty=1,lwd=2.5,col=c("blue"),bty="n")
#to add your specific axis format to the x-axis
axis.POSIXct(side=1, at=tickpos, format="%Y-%b")
```
####DHW calculations from full dataset####
```{r}
#subsetting df to create climatology avoiding bleaching years
nbtemp<-subset(temp, year !='2005' & year !='2014')

#daily calcs from nb years to find MMM
daily<-data.frame("DayRange"=tapply(nbtemp$Temp, nbtemp$d, function(x) range(x)[2]-range(x)[1]),"DayMin"=tapply(nbtemp$Temp, nbtemp$d, min),"DayMax"=tapply(nbtemp$Temp, nbtemp$d, max), "DayMean"=tapply(nbtemp$Temp, nbtemp$d, mean))
daily$Date<-strptime(row.names(daily),format="%d_%m_%Y")
str(daily)
daily<-daily[order(daily$Date),]
daily$month<-format(daily$Date, "%B-%Y")
monthlymean<-data.frame("MonthlyMeans"=tapply(daily$DayMean,daily$month, mean))
monthlymean$month<-row.names(monthlymean)
monthlymean$year<-format(strptime(paste0("01-",monthlymean$month),format="%d-%B-%Y"),"%Y")
str(monthlymean)
monthlymean$mo<-format(strptime(paste0("01-",monthlymean$month),format="%d-%B-%Y"),"%B")
Summarize(MonthlyMeans~mo, data=monthlymean, digits=3)
#MMM = 31.173 August nb
```

#Climatology = maximum of the monthly mean SST climatology, mean temperature of the climatologically warmest month at the location
#Hotspot = difference between a nighttime SST value and the corresponding climatology value
```{r}
# df to calculate Hotspots and DHW DHD
tempdaily$month<-format(tempdaily$date, "%B-%Y")
monthlymean<-data.frame("MonthlyMeans"=tapply(tempdaily$dayMean,tempdaily$month, mean))
monthlymean$month<-row.names(monthlymean)
monthlymean$year<-format(strptime(paste0("01-",monthlymean$month),format="%d-%B-%Y"),"%Y")
head(monthlymean)
monthlymean$mo<-format(strptime(paste0("01-",monthlymean$month),format="%d-%B-%Y"),"%B")
Summarize(MonthlyMeans~mo, data=monthlymean, digits=3)
#MMM = 30.955 August .....weird, lower than when bleaching years kept
```

#calculate hotspots and DHW
```{r}
#if/else wont work with NA remove
tempdaily<-na.omit(tempdaily)
Hotspots<-NULL
for(i in 1:nrow(tempdaily)){
  if((tempdaily[i,"dayMean"]-30.955)>=1){
    z<-tempdaily[i,"dayMean"]-30.955
  }else{
    z<-0
  }
  Hotspots<-c(Hotspots,z)
}

tempdaily$hotspot<-Hotspots
#DHW = the accumulation of HotSpots at that location over a rolling 12-week time period,84 is no of days
tempdaily$DHW<-(1/7)*c(rep(0,83),rollapply(tempdaily$hotspot,84,sum))
tempdaily$DHD<-c(rep(0,83),rollapply(tempdaily$hotspot,84,sum))
headTail(tempdaily)
tempdaily$year<-format(strptime(paste0("01-",tempdaily$month),format="%d-%B-%Y"),"%Y")
```
#monthly stats
```{r}
monthly<-data.frame("MonthlyMin"=tapply(tempdaily$dayMin,tempdaily$month, min),"MonthlyMax"=tapply(tempdaily$dayMax,tempdaily$month, max),"MonthlyRange"=tapply(tempdaily$dayRange, tempdaily$month, max), "MonthlyMean"=tapply(tempdaily$dayMean,tempdaily$month, mean), "MonthlyDHW"=tapply(tempdaily$DHW ,tempdaily$month, max),"MonthlyDHD"=tapply(tempdaily$DHD,tempdaily$month, max))
write.table(monthly,file="2005-2017_NHF_monthlyclimatology.txt",sep="\t")
```
#plot climatology (SST and DHW) 
```{r}
par(mar=c(5,5,3,5))
plot(tempdaily$date,tempdaily$dayMean, type="l", lwd= 2, ylab="", xlab="Date", xaxt='n',ylim=c(18,34), col="blue")
abline(h=29.7,lty=3,lwd=2, col="black")
legend("topright",c("Temp","DHW"),lty=c(1,2),lwd=2.5,col=c("blue","red"),bty="n")
#add in DHW with separate yaxis
par(new=T)
plot(tempdaily$date,tempdaily[,"DHW"],type="l",lty=2, col="red", lwd=3, xaxt='n',yaxt='n',xlab='',ylab='', ylim=c(0,10))
axis(side=4, cex.axis=1)
mtext("DHW (°C week)",cex=1.5,side=4,line=3)
mtext("Mean in-situ Temperature (°C)",cex=1.5,side=2,line=3)
#to add your specific axis format to the x-axis
tickpos<-seq(as.POSIXct("2005-06-04"),as.POSIXct("2021-07-30"),by="1 year")
axis.POSIXct(side=1, at=tickpos, format="%Y-%b")
```
##calculate DR90 and days over threshold##
```{r}
#calculate DR90%, the 90th quartile for daily temp ranges over rolling 10day
head(tempdaily)
DR90<-rollapply(tempdaily$dayRange,width=10,FUN="quantile",p=0.9,by=10)

#loop over number of rows by date to count number of days over 30.5
#also have to copy values for each row
monthly$daysBT <- tapply(tempdaily$dayMax, tempdaily$month, function(x) sum(x > 30.5))
monthly$days31 <- tapply(tempdaily$dayMax, tempdaily$month, function(x) sum(x > 31))
monthly$days32 <- tapply(tempdaily$dayMax, tempdaily$month, function(x) sum(x > 32))
monthly$days33 <- tapply(tempdaily$dayMax, tempdaily$month, function(x) sum(x > 33))
monthly$date<-row.names(monthly)
monthly$year<-format(strptime(paste0("01-",monthly$date),format="%d-%B-%Y"),"%Y")
monthly$month<-format(strptime(paste0("01-",monthly$date),format="%d-%B-%Y"),"%B")
monthly$month <- factor(monthly$month, levels = month.name)
monthly$year<-as.factor(monthly$year)
```
##plotting monthly summary of mean, max, dtr, and days over BT ##
```{r}
#remove 2016 bc incomplete
monthly2<-subset(monthly, year !='2016')

# Define the number of colors you want
nb.cols <- 16
#cols<-scale_color_brewer(palette = "Spectral",direction=-1)
mycolors <- colorRampPalette(brewer.pal(11, "Spectral"))(nb.cols)

#ggplot points 
p1<-ggplot(monthly2,aes(x=month, y=MonthlyMax, group=year)) + geom_line(aes(color=year),size=1.5) +  scale_color_manual(values = mycolors) +              
  ylab("Max (°C)") + ggtitle("A") + theme_bw(base_size=12) +theme(axis.title.x=element_blank()) + scale_x_discrete(limits = month.name)

p2<-ggplot(monthly2,aes(x=month, y=MonthlyRange, group=year)) + geom_line(aes(color=year),size=1.5,show.legend = FALSE)+  scale_color_manual(values = mycolors) +              
  ylab("Range (°C)") + ggtitle("B") + theme_bw(base_size=12) +theme(axis.title.x=element_blank()) + scale_x_discrete(limits = month.name) 

p3<-ggplot(monthly2,aes(x=month, y=MonthlyMean,group=year))+ geom_line(aes(color=year),size=1.5,show.legend = FALSE)+ scale_color_manual(values = mycolors) +       
  ylab("Mean (°C)") + ggtitle("C")  + theme_bw(base_size=12) +theme(axis.title.x=element_blank()) + scale_x_discrete(limits = month.name) 

p4<-ggplot(monthly2,aes(x=month, y=daysBT, group=year))+ geom_line(aes(color=year),size=1.5,show.legend = FALSE) + scale_color_manual(values = mycolors) +       
  ylab("Days > BT (30.5 °C)") + ggtitle("D")  + theme_bw(base_size=12) +theme(axis.title.x=element_blank()) + scale_x_discrete(limits = month.name) 

prow <- plot_grid(
  p1 + theme(legend.position="none"),
  p2 + theme(legend.position="none"),
  p3 + theme(legend.position="none"),
  p4 + theme(legend.position = 'none'),
  align = 'vh',
  hjust = -1,
  nrow = 4
)
# extract a legend that is laid out horizontally
legend_b <- get_legend(
  p1 + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom")
)

# add the legend underneath the row we made earlier. Give it 20%
# of the height of one plot (via rel_heights).
plot_grid(prow, legend_b, ncol = 1, rel_heights = c(1, .1))
```
